---
title: "Data Wells: Technical file"
author: 
  - "Nathan Alexander^[Howard University]"
  - "Kade Davis^[Morehouse College]"
  - "Quantitative Histories Workshop"
output:   
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true  # Enable floating TOC in the sidebar
    number_sections: true
    theme: cerulean
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(here)
here::i_am("data-wells.Rmd")
```

# Overview

We provide here the technical code and methods to accompany our analysis of data on lynching and policing. We examine some of the historical and contemporary contexts of state violence and social control using a media frames analysis regarding the state's means of production towards social control -- we then situate this idea using Goodwin's (1992) analysis around *professional visions* and we broadly frame our analysis within the research in Black and African American studies. We also consider some of the structures, assumptions, and quantitative models related to the analysis of the historical data. We make use of original source data from Ida B. Wells-Barnett's *The Red Record* and the Washington Post *Fatal Force* database.

# Data

## The Washington Post Fatal Force database

```{r}
# fatal database
fatal <- read.csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/refs/heads/master/v2/fatal-police-shootings-data.csv")
str(fatal) 
```

```{r}
# fix vars
# change vars to more appropriate formats
fatal$date <- as.Date(fatal$date) # check/change to date format

fatal$year <- as.numeric(format(fatal$date, "%Y"))

fatal$age <- as.numeric(fatal$age)

fatal$gender[fatal$gender == ""] <- NA
fatal$gender <- as.factor(fatal$gender)
fatal$gender <- droplevels(fatal$gender)

fatal <- fatal %>%
  mutate(
    race_category = case_when(
      race == ""       ~ NA_character_,
      race == "A"      ~ "Asian",
      race == "B"      ~ "Black",
      race == "H"      ~ "Hispanic",
      race == "N"      ~ "Native American",
      race == "O"      ~ "Other",
      race == "W"      ~ "White",
      race == "B;H"    ~ "Black, Hispanic",
      race == "N;H"    ~ "Native, Hispanic",
      race == "W;A"    ~ "White, Asian",
      race == "W;B"    ~ "White, Black",
      race == "W;B;N"  ~ "White, Black, Native",
      race == "W;H"    ~ "White, Hispanic",
      TRUE             ~ "Other"
    )
  )
fatal$race_category <- as.factor(fatal$race_category)

library(dplyr)

fatal <- fatal %>%
  mutate(
    black = case_when(
      race == "B" ~ "Black",
      grepl(";", race) & grepl("B", race) ~ "Black Other",  # Multiracial with Black
      TRUE ~ "Non-Black"
    )
  )

# convert to factor for ordered levels
fatal$black <- factor(fatal$black, 
                                 levels = c("Black", "Black Other", "Non-Black"))

fatal$was_mental_illness_related <- as.logical(fatal$was_mental_illness_related)

fatal$body_camera <- as.logical(fatal$body_camera)

str(fatal)

# view a summary of the data
summary(fatal)

# create a two-column transfer df to match state to abb
transfer <- tibble(state = state.name) %>% 
  tibble(abb = state.abb) %>% 
  bind_rows(tibble(state = "District Of Columbia", abb = "DC")) # add DC
transfer
tail(transfer)

# add a state name variable to the fatal df
fatal$state.name <- state.name[match(fatal$state, transfer$abb)]
fatal %>% 
  mutate(state.abb = state) %>% 
  relocate(id, date, state.name, state.abb) -> fatal

# create a year column
# format to 20YY
fatal.year <- format(fatal$date, format="20%y") 
fatal$year <- fatal.year # add column to df
fatal$year <- as.numeric(fatal$year)
fatal %>% relocate(id, date, year, state.name, state.abb) -> fatal
tail(fatal)
```

Subsetting data for 2023 and 2024.

```{r}
df2023 <- fatal %>% 
  filter(year == 2023)

df2024 <- fatal %>% 
  filter(year == 2024)
```

## Ida B. Wells-Barnett's The Red Record data

We then load data from The Red Record. Data are gathered from two sources. We conduct a set of cross references to confirm the final selection of case studies for the analysis.

```{r}
# data from forked repo on IdaBWellsProject
df1893 <- read.csv("https://raw.githubusercontent.com/quant-shop/IdaBWellsProject/master/RedRecord/redrecord1893.csv")

df1894 <- read.csv("https://raw.githubusercontent.com/quant-shop/IdaBWellsProject/master/RedRecord/redrecord1894.csv")
```

```{r}
# records from quant shop entry
df1892 <- read_csv("../data/Red Record Lynching Record - 1892.csv")

df1893b <- read_csv("../data/Red Record Lynching Record - 1893.csv")

df1894b <- read_csv("../data/Red Record Lynching Record - 1893.csv")
```

## Standardizing data frames

```{r}
# standardize data frames
# --- Fix date columns for 1893 and 1894 ---
df1893$Date <- as.Date(df1893$Date, format="%b. %d, %Y")
df1894$Date <- as.Date(df1894$Date, format="%b. %d, %Y")

# when months are not abbreviated with a dot, use this alternative:
# df1893$Date <- as.Date(df1893$Date, format="%b %d, %Y")
# df1894$Date <- as.Date(df1894$Date, format="%b %d, %Y")

# --- standardize column names for 1893 and 1894 ---
names(df1893) <- c("name", "location", "date", "alleged_crime", "latitude", "longitude")
names(df1894) <- c("name", "location", "date", "alleged_crime", "latitude", "longitude")

# --- check and fix types in 2023 and 2024 ---
# factorize categorical variables if needed:
factor_cols <- c("state.name", "state.abb", "threat_type", "flee_status", "armed_with",
                 "city", "county", "state", "gender", "race", "race_source", "location_precision")

df2023[factor_cols] <- lapply(df2023[factor_cols], as.factor)
df2024[factor_cols] <- lapply(df2024[factor_cols], as.factor)

# --- add a 'year' column to 1893 and 1894 for consistency ---
df1893$year <- 1893
df1894$year <- 1894

# --- add missing columns to 1893/1894 with NA values for merging ---
missing_cols <- setdiff(names(df2023), names(df1893))
for(col in missing_cols) df1893[[col]] <- NA
df1893 <- df1893[, names(df2023)] # Reorder columns

missing_cols <- setdiff(names(df2023), names(df1894))
for(col in missing_cols) df1894[[col]] <- NA
df1894 <- df1894[, names(df2023)]

all_years <- rbind(df1893, df1894, df2023, df2024)
```


```{r, include=F, eval=F}
write.csv(df1892, "../data/df1892.csv", row.names = FALSE)
write.csv(df1893, "../data/df1893.csv", row.names = FALSE)
write.csv(df1893b, "../data/df1893b.csv", row.names = FALSE)
write.csv(df1894, "../data/df1894.csv", row.names = FALSE)
write.csv(df1894b, "../data/df1894b.csv", row.names = FALSE)
write.csv(df2023, "../data/df2023.csv", row.names = FALSE)
write.csv(df2024, "../data/df2024.csv", row.names = FALSE)
```

# Maps

```{r}
library(ggplot2)
library(maps)
#install.packages("maps")

# get map data for US states
states_map <- map_data("state")
```

## Fatal Force

```{r}
# lowercase to match maps of state names
fatal <- fatal %>%
  mutate(region = tolower(state))  # or whatever your state column is called
```

```{r}
# states to higlight from wells 1892 data - top five states
states1892 <- c(
  "louisiana", 
  "tennessee", 
  "arkansas",
  "alabama",
  "georgia",
  "mississippi",
  "texas",
  "florida",
  "kentucky",
  "wyoming"
  )
states_map_1892 <- states_map %>%
  mutate(highlight = ifelse(region %in% states1892, "highlight", "normal"))
```

Add dots for each fatal incident.

```{r}
ggplot() +
  geom_polygon(
    data = states_map_1892,
    aes(x = long, y = lat, group = group, fill = highlight),
    color = "white", size = 0.2
  ) +
  scale_fill_manual(values = c("highlight" = "red", "normal" = "grey80")) +
  geom_point(
    data = fatal,
    aes(x = longitude, y = latitude),
    color = "blue", alpha = 0.6, size = 2
  ) +
  coord_fixed(1.3) +
  theme_void() +
  labs(title = "US Map with Highlighted States and Incident Dots")
```

### Black men in 2022

```{r}
fatal_black_men_2022 <- fatal %>%
  filter(
    race_category == "Black" | race == "B",
    gender == "male" | gender == "M", # Adjust based on your gender coding
    year == 2022
  )

fatal_black_men_2022 <- fatal_black_men_2022 %>%
  mutate(state_lower = tolower(state)) 
```

```{r}
ggplot() +
  geom_polygon(
    data = states_map,
    aes(x = long, y = lat, group = group),
    fill = "gray90",
    color = "white"
  ) +
  geom_point(
    data = fatal_black_men_2022,
    aes(x = longitude, y = latitude),
    color = "black",
    alpha = 0.7,
    size = 2
  ) +
  coord_fixed(1.3) +
  theme_void() +
  labs(
    title = "Fatal Police Shootings of Black Men in the US, 2022",
    caption = "Data source: Washington Post Fatal Force Database"
  )

```

```{r}
states_map_1892 <- states_map_1892 %>%
  mutate(
    highlight = ifelse(region %in% states1892, "highlight", "normal")
  )
```

## Integreted maps

Here, we integrate the fatal force 2022 data for Black men with the 1892 records.

```{r}
ggplot() +
  geom_polygon(
    data = states_map_1892,
    aes(x = long, y = lat, group = group, fill = highlight),
    color = "white", size = 0.2
  ) +
  scale_fill_manual(
    values = c("highlight" = "red", "normal" = "gray80"),
    guide = "none"
  ) +
  coord_fixed(1.3) +
  theme_void() +
  labs(
    title = "Top 10 States, Lynching in 1892"
  )

```



```{r}
ggplot() +
  geom_polygon(
    data = states_map_1892,
    aes(x = long, y = lat, group = group, fill = highlight),
    color = "white", size = 0.2
  ) +
  scale_fill_manual(
    values = c("highlight" = "red", "normal" = "gray80"),
    guide = "none"
  ) +
  geom_point(
    data = fatal_black_men_2022,
    aes(x = longitude, y = latitude),
    color = "black", alpha = 0.7, size = 2
  ) +
  coord_fixed(1.3) +
  theme_void() +
  labs(
    title = "Top 10 States - Lynching, 1892 (red); Black Male Fatal Shootings, 2022 (dots)"
  )


```

# Tables

Tables without context.

Flee status.

```{r}
# Frequency table
table(fatal$flee_status)

# Proportion table (relative frequencies)
prop.table(table(fatal$flee_status))
```
Armed with.

```{r}
# Frequency table
table(fatal$armed_with)

# Proportion table (relative frequencies)
prop.table(table(fatal$armed_with))
```

```{r}
#kable table
# Create a frequency table
armed_tbl <- as.data.frame(table(fatal$armed_with))
colnames(armed_tbl) <- c("armed_with", "count")

# Add classification
armed_tbl <- armed_tbl %>%
  mutate(category_type = case_when(
    grepl(";", armed_with) ~ "Combined",
    armed_with %in% c("unknown", "undetermined", "not_reported", "NA", "") ~ "Missing/Ambiguous",
    TRUE ~ "Clear"
  ))

# Add % column (optional)
armed_tbl <- armed_tbl %>%
  mutate(proportion = round(count / sum(count), 4))

# Sort by count descending
armed_tbl <- armed_tbl %>%
  arrange(desc(count))

# Print as kable
kable(armed_tbl, caption = "Summary of `armed_with` Categories in `fatal` Dataset")

```



Additional context.

```{r}

# define race categories to *keep* (single, mutually exclusive)
keep_race <- c("B", "W", "H")

# define armed statuses to keep 
keep_armed <- c("gun", "unarmed", "knife") # Add others as appropriate; avoid "multiple", "unknown"

fatal_single <- fatal %>%
  filter(race %in% keep_race, armed_with %in% keep_armed)

# Flee Status x Race
table(fatal_single$flee_status, fatal_single$race)
# Armed With x Race
table(fatal_single$armed_with, fatal_single$race)
# Flee Status x Armed With x Race
ftable(fatal_single$flee_status, fatal_single$armed_with, fatal_single$race)


race_labels <- c("B" = "Black", "W" = "White", "H" = "Hispanic")
fatal$race <- recode(fatal$race, !!!race_labels)

```

A better example.

```{r}
# we can also use the census dat to make sense of the proportion of the population for a given county

```




